Improving Haskell Theory Exploration

### 2015-12-09 TransferReport/related.tex
Finished feature extraction/machine learning

One major difficulty with formal mathematics as a domain in which to apply statistical machine learning is the use of \emph{structure} to encode information in objects. In particular, \emph{trees} appear in many places: from inductive datatypes, to recursive function definitions; from theorem statements, to proof objects. Such nested structures may extend to arbitrary depth, which makes them difficult to process with many machine learning algorithms. The solution, as described in \S \ref{sec:featureextraction}, is to use \emph{feature extraction}. However, our method is not the only possible way to encode recursive structures as fixed-size features.

The simplest way to encode such inputs is to simply choose a desired size, then pad anything smaller and truncate anything larger. We use this to make our matrices a uniform size, borrowing the idea from ML4PG. Care must be taken to ensure that we are not discarding too much information, that we are not producing features with too many dimensions to be practical, and that there is a uniform ``meaning'' to each feature across different feature vectors. In our case, we transform the recursive structure of expressions into matrices in order to give meanings such as ``the $i$th token at the $j$th level of nesting''.

Truncation works best when the input data contains no redundancy and is arranged with the most significant data first (in a sense, it is ``big-endian''). The less these assumptions hold, the less we can truncate. These seem to hold in the case of Haskell expressions, since the higher levels of the syntax tree are the most semantically significant; for example, due to laziness, the lower levels may never even be evaluated.

@@

There are some machine learning algorithms which can handle variable-size input. In particular, \emph{recurrent artificial neural networks} (RNNs) can learn \emph{sequences}, such as audio or written sentences. For example, we could render our expressions into a linear sequence of tokens (e.g. s-expressions) and process these with an RNN. Unfortunately there are several problems with this approach: firstly, we must train RNNs in a \emph{supervised} way, unlike clustering which is \emph{unsupervised}; this requires us to define an error signal which the RNN should minimise. There are ways to achieve clustering this way, for example by having the RNN \emph{compress} each expression to a fixed size we get a dimension reduction technique. \iffalse and, in fact, clustering CITE Clustering by compression\fi In this case the error signal is the difference between the original input and the output after decompression. This is the basis of \emph{autoencoders}.

Another problem with RNNs is the difficulty of training them. If we extend the standard backpropagation algorithm (used for non-recurrent neural networks) to handle the cycles present in RNNs, we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}. However, this suffers a problem known as the \emph{vanishing} or \emph{exploding gradient}: error values grow or decay exponentially as they propagate back through the cycles, which prevents effective learning of correlations across a sequence, undermining the main advantage of RNNs. The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques to train, rather than backpropagation) and \emph{long short-term memory} (LSTM; introducing a few special, nodes to persist values for long time periods, unaffected by training \citep{hochreiter1997long}).

The application of \emph{kernel methods} to structured information is discussed in \citep{Gartner2003}, where the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size.

% TODO
\citep{Gartner2003}
\citep{Oveisi.Oveisi.Erfanian.ea:2012}
\citep{bakir2007predicting}
\citep{conf/ijcai/Plate91}
\citep{goller1996learning}
\citep{kwasny1995tail}
\citep{pollack1990recursive}
\citep{zanzotto2012distributed}

\iffalse

Machine learning over structured data:
1D is common: parsing natural language
2D is common; images
Trees are fractal
Backpropagation through structure
LSTM with recursive structure
Most work tries to identify structure; we already have it

Recurrent neural networks
Backpropagation through structure

>>>>

One major difficulty when applying statistical machine learning algorithms to \emph{languages}, such as Haskell, is the appearance of recursive structures. This can lead to nesting to arbitrary depths, which makes such expressions difficult to compare in numerical ways. The solution, as described in \S \ref{sec:featureextraction}, is to use \emph{feature extraction}; however, our method is not the only possible way to encode recursive structures as fixed-size features.

The simplest way to encode such inputs is to simply choose a desired size, then pad anything smaller and truncate anything larger. We use this to make our matrices a uniform size, borrowing the idea from ML4PG. Care must be taken to ensure that we are not discarding too much information, that we are not producing features with too many dimensions to be practical, and that there is a uniform ``meaning'' to each feature across different feature vectors. In our case, we avoid many of these problems by transforming the recursive structure of expressions into matrices first; this gives each feature a stable meaning such as ``the $i$th token from the left at the $j$th level of nesting''.

Truncation works best when the input data is arranged with the most significant data first (in a sense, it is ``big-endian''). This is the case for Haskell expressions, since the higher levels of the syntax tree are the most semantically significant; for example, the lower levels may never even be evaluated due to laziness. This allows us to truncate more aggressively than if the leaves were most significant.

@@

There are machine learning algorithms which can handle variable-size input, but these are often \emph{supervised} algorithms which require an externally-provided error function to minimise. Error functions can be given for clustering, for example k-means implicitly minimises the function given in equation \ref{eq:kmeansobjective}, but unsupervised algorithms may be preferred for efficiency as they are more direct.

One example of learning from variable-size input is to use \emph{recurrent artificial neural networks} (RNNs). These contain cyclic connections between neurons, unlike the traditional acyclic ``feed-forward'' ANNs, allowing state to persist between observations. In this way, each data point can be divided into a sequence of arbitrary length, for example a linear sequence of tokens like an s-expression, and fed into an RNN for processing.

Unfortunately RNNs are difficult to train. The standard way to train ANNs is the back-propagation algorithm; when this is extended to handle cycles we get the \emph{backpropagation through time} algorithm \citep{werbos1990backpropagation}. However, this suffers a problem known as the \emph{vanishing} (or \emph{exploding}) \emph{gradient}: error values change exponentially as they propagate back through the cycles, which prevents effective learning of correlations across a sequence, undermining the main advantage of RNNs. The vanishing gradient problem is the subject of current research, with countermeasures including \emph{neuroevolution} (using evolutionary computation techniques rather than back-propagation) and \emph{long short-term memory} (LSTM; introducing special nodes to ``store'' state, rather than having them loop around a cycle \citep{hochreiter1997long}).

Using sequences to represent recursive structures is also problematic: if we want our learning algorithm to exploit structure (such as the depth of a token), it will have to discover how to parse the sequences for itself, which seems wasteful. The \emph{back-propagation through structure} approach \citep{goller1996learning} is a more direct solution to this problem, using a feed-forward ANN to learn recursive distributed representations \citep{pollack1990recursive} which correspond exactly to the recursive structure of the inputs. Such distributed representations can also be used for sequences, which is useful for encoding trees with a non-uniform branching factor \citep{kwasny1995tail}. More recent work has investigated storing recursive structures inside LSTM cells \citep{zhu2015long}.

A simpler alternative for generating recursive distributed representations is to use circular convolution \citep{conf/ijcai/Plate91}. Although promising results are shown for its use in \emph{distributed tree kernels} \citep{zanzotto2012distributed}, our preliminary experiments in applying circular convolution to functional programming expressions found most of the information to be lost in the process; presumably as the expressions are too small.

\emph{Kernel methods} have also been applied to structured information, for example in \citep{Gartner2003} the input data (including sequences, trees and graphs) are represented using \emph{generative models}, such as hidden Markov models, of a fixed size suitable for learning. Many more applications of kernel methods to structured domains are given in \citep{bakir2007predicting}.
